---
title: "DS105M 2022 Week08"
output: html_notebook
---

------------------------------------------------------------------------

**DS105M 2022 Week08**

**Topic:** (Re-)shaping data, data normalisation & databases

**Author:** [\@jonjoncardoso](github.com/jonjoncardoso)

**Date:** 15 November 2022

------------------------------------------------------------------------

# Setup

You might need to `install.packages()` if you do not have the libraries below.

**💡 Tip:** Switch to the "Visual" Mode instead of "Source" at the top of this screen.

# Import Libraries

```{r}
library(tidyverse) #rvest is part of tidyverse
```

# Data Collection

**Web Scraping:** [Frequent 'R' Questions](https://stackoverflow.com/questions/tagged/R)

```{r}
# read the content of the site
response_html <- read_html("https://stackoverflow.com/questions/tagged/R")
```

The `response_html` object returned by the `rvest::read_html` function is not as easy to explore as the one we get with the Python's BeautifulSoup version of this tutorial. But they encode the same info.

You can get a sense for it by printing the object:

```{r}
print(response_html)
```

Keep a tab with the documentation of rvest to consult how to do things: [rvest documentation](https://rvest.tidyverse.org/).

```{r}
phonetic_word <- if(...){
  "-"
}else{
  getting_phonetic_json$phonetics
}
```

## What will we extract?

Here we care about extracting all the posts:

-   Title
-   Date
-   Summary statistics (number of votes, answers and views)

Let's explore how to do that step-by-step.

# Step-by-Step is the way to go

💡 Here's the key takeaway of this session: don't write code to collect all the questions. Write code to collect **ONLY ONE** question first.

## Extracting the right `div`s

After exploring the HTML, I found out I want all the `<div class="s-post-summary">` inside the main `<div id="questions">`:

```{r}
all_posts <-  
  html_element(response_html, css="div#questions") %>% 
  html_elements(css="div.s-post-summary")

cat(paste0("We found ", length(all_posts) , " posts")) # cat is another way to `print()` stuff
```

## What is inside each `div`?

```{r}
current_post <- all_posts[[1]] # Select just the first one
current_post %>% xml2::html_structure()
```

Another way to explore this data more visually is to save it and open it on a code editor such as [VScode](https://code.visualstudio.com/download) (We used it in the [💻 Week 07 - Lab Roadmap](https://lse-dsi.github.io/lse-ds105-course-notes/weeks/week07/lab.html))

## Let's break it all down to get the summary statistics

```{r}
current_post %>% html_elements(css="div.s-post-summary--stats-item")
```

There are two ways to get the info I want:

-   I could extract the element `title` inside of the `<div>`
-   I could extract the text inside the `<span>` elements.

**⁉️ Still puzzled about what some HTML element tags represent? The book *Duckett, Jon. HTML and CSS: Design and Build Websites. New York: Wiley, 2011.* is a great resource.**

Let's do it the hard way, extract from `<span>`. I will show how to approach this with a `for` loop first, since this is probably how you learned how to do things with programming.

If I just want to print stuff, I could just navigate the tree:

```{r}
current_div_summary <- current_post %>% html_elements(css="div.s-post-summary--stats-item")
current_div_summary <- current_div_summary[[1]]
current_div_summary
```

Let's get the info we want:

```{r}
current_div_summary %>% html_element("span.s-post-summary--stats-item-unit") %>% html_text()
```

Let's get the info we want:

```{r}
current_div_summary %>% html_element("span.s-post-summary--stats-item-number") %>% html_text()
```

This would be better if we saved it in a **named list**:

```{r}
one_summary_dict <- 
  data.frame(unit=current_div_summary %>% html_element("span.s-post-summary--stats-item-unit") %>% html_text(),
      number=current_div_summary %>% html_element("span.s-post-summary--stats-item-number") %>% html_text()    
)
one_summary_dict
```

## Putting the pieces of the puzzle together: gathering summary statistics for a single post

However, the part people struggle the most is figuring out how to **pre-process and store the data** correctly.

If you follow the strategy above, it will be much easier to think about that. Let's explore together how we would gather **all statistics** from a `current_div_summary`:

```{r}
all_stats = data.frame() # an empty dataframe that will be filled with multiple named lists (do you see why? ask your lecturer if not)

all_div_summaries <- current_post %>% html_elements("div.s-post-summary--stats-item")

for(j  in 1:length(all_div_summaries)){
  
  current_div_summary <- all_div_summaries[[j]]
  
  one_summary_dict <- 
    data.frame(unit=current_div_summary %>% 
              html_element("span.s-post-summary--stats-item-unit") %>% 
              html_text(),
        number=current_div_summary %>% 
               html_element("span.s-post-summary--stats-item-number") %>% 
               html_text()    
        )
  
  all_stats <- bind_rows(all_stats, one_summary_dict)
}

# What is in `all_stats` now?
all_stats
```

**🎯 ACTION POINT**: Would you organise this dataframe differently?

> your text here

## Going one step up: gathering summary statistics for ALL posts

How would we gather the summary for **ALL posts**?

-   Time to learn about `bind_rows`

```{r}

all_stats = data.frame() # an empty dataframe that will be filled with multiple named lists (do you see why? ask your lecturer if not)

for(i in 1:length(all_posts)){

  current_post <- all_posts[[i]]
  
  all_div_summaries <- current_post %>% html_elements("div.s-post-summary--stats-item")

  for(i  in 1:length(all_div_summaries)){
    
    current_div_summary <- all_div_summaries[[i]]
    
    one_summary_dict <- 
      data.frame(unit=current_div_summary %>% 
                html_element("span.s-post-summary--stats-item-unit") %>% 
                html_text(),
          number=current_div_summary %>% 
                 html_element("span.s-post-summary--stats-item-number") %>% 
                 html_text()    
          )
    
    all_stats <- bind_rows(all_stats, one_summary_dict)
  }
}

all_stats
```

**🤔 Q: What is the problem with the data frame above?**

A: Well, I have the statistics about all the 15 posts but I don't know which one is which!

Let's fix that:

-   Time to learn about [str_trim](https://stringr.tidyverse.org/reference/str_trim.html) (also a tidyverse function)

```{r}

all_stats = data.frame() 

for(i in 1:length(all_posts)){

  current_post <- all_posts[[i]]
  post_title <- 
      current_post %>% 
      html_element("h3.s-post-summary--content-title") %>%  
      html_text() %>% 
      str_trim()
  
  # Notice anything different in the html_element here? 
  # Raise your hand and call your lecture if you don't see any difference.
  post_time  <- 
        current_post %>% 
      html_element("time span") %>%  
      html_attr("title")
  
  all_div_summaries <- current_post %>% html_elements("div.s-post-summary--stats-item")

  for(i  in 1:length(all_div_summaries)){
    
    current_div_summary <- all_div_summaries[[i]]
    
    one_summary_dict <- 
      data.frame(unit=current_div_summary %>% 
                html_element("span.s-post-summary--stats-item-unit") %>% 
                html_text(),
          number=current_div_summary %>% 
                 html_element("span.s-post-summary--stats-item-number") %>% 
                 html_text() ,
          post_title=post_title,
          post_time=post_time
          )
    
    all_stats <- bind_rows(all_stats, one_summary_dict)
  }
}

all_stats
```

# Functions

Don't underestimate the power of functions!

In fact, you might find it easier to work with them than with `for` loops.

## A function just to parse the `current_div_summary`

Check [roxygen2 convention](https://cran.r-project.org/web/packages/roxygen2/vignettes/rd.html) for documenting functions

```{r}
#' 
#' Let's write documentation together
#'
parse_div_summary <- function(current_div_summary){
    one_summary_dict <- 
      data.frame(unit=current_div_summary %>% 
                html_element("span.s-post-summary--stats-item-unit") %>% 
                html_text(),
          number=current_div_summary %>% 
                 html_element("span.s-post-summary--stats-item-number") %>% 
                 html_text()    
          )
    one_summary_dict
}

```

```{r}
parse_div_summary(current_div_summary)
```

## A function to parse all summaries from a `current_post`

Time to learn about `lapply`

```{r}
parse_all_summaries <- function(current_post){
  
  all_div_summaries <- current_post %>% html_elements("div.s-post-summary--stats-item")
  
  all_summaries <- lapply(all_div_summaries, parse_div_summary)
  all_summaries <- bind_rows(all_summaries) # R's equivalent to python's pd.concat
  all_summaries 
}

```

```{r}
parse_all_summaries(current_post)
```

## A function to parse other info about a `current_post`

```{r}
parse_post_div <- function(current_post){
    
  post_title <- 
      current_post %>% 
      html_element("h3.s-post-summary--content-title") %>%  
      html_text() %>% 
      str_trim()
  post_time  <- 
        current_post %>% 
      html_element("time span") %>%  
      html_attr("title")
    
  df_summary <- parse_all_summaries(current_post)
  
  # We can reuse the dataframe that returned from `parse_all_summaries`
  df_summary$post_title <- post_title
  df_summary$post_time <- post_time
  
  df_summary 
  
}

```

```{r}
parse_post_div(current_post)
```

## A function to parse ALL posts

Instead of growing a list using `for` loops, let's use list comprehension and use `pd.concat()`

Time to learn about `reset_index()` and `inplace=True`

```{r}
parse_all_posts <- function(response_html){
    
  all_posts <-  
    html_element(response_html, css="div#questions") %>% 
    html_elements(css="div.s-post-summary")
  
  df_results <- bind_rows(lapply(all_posts, parse_post_div))
    
  # Optional: If you are also annoyed at the zero indices
  # df_results.reset_index(drop=True, inplace=True)
  df_results 
  
}

```

```{r}
parse_all_posts(response_html)
```

## Putting it all together

-   Use RMarkdown for prototyping, but not for reusing code
-   Store your functions in a separate file

```{r}
parse_div_summary <- function(current_div_summary){
    one_summary_dict <- 
      data.frame(unit=current_div_summary %>% 
                html_element("span.s-post-summary--stats-item-unit") %>% 
                html_text(),
          number=current_div_summary %>% 
                 html_element("span.s-post-summary--stats-item-number") %>% 
                 html_text()    
          )
    one_summary_dict
}

parse_all_summaries <- function(current_post){
  
  all_div_summaries <- current_post %>% html_elements("div.s-post-summary--stats-item")
  
  all_summaries <- lapply(all_div_summaries, parse_div_summary)
  all_summaries <- bind_rows(all_summaries) # R's equivalent to python's pd.concat
  all_summaries 
}

parse_post_div <- function(current_post){
    
  post_title <- 
      current_post %>% 
      html_element("h3.s-post-summary--content-title") %>%  
      html_text() %>% 
      str_trim()
  post_time  <- 
        current_post %>% 
      html_element("time span") %>%  
      html_attr("title")
    
  df_summary <- parse_all_summaries(current_post)
  
  # We can reuse the dataframe that returned from `parse_all_summaries`
  df_summary$post_title <- post_title
  df_summary$post_time <- post_time
  
  df_summary 
  
}

parse_all_posts <- function(response_html){
    
  all_posts <-  
    html_element(response_html, css="div#questions") %>% 
    html_elements(css="div.s-post-summary")
  
  df_results <- bind_rows(lapply(all_posts, parse_post_div))
    
  # Optional: If you are also annoyed at the zero indices
  # df_results.reset_index(drop=True, inplace=True)
  df_results 
  
}
```

**🎯 ACTION POINT**:

-   Clear the environment:

```{r}
rm(list=ls())
```

Read more about the `rm` function like [here](https://stackoverflow.com/a/43673112/843365)

-   In the same folder as this notebook, create a new file `stackparser_questions.R`

The code below should return the same `df` as we have been creating:

```{r}
source("stackparser_questions.R")
response_html <- read_html("https://stackoverflow.com/questions/tagged/R")
df <- parse_all_posts(response_html)
head(df)
```

### Resulting DataFrame

```{r}
response_html <- read_html("https://stackoverflow.com/questions/tagged/R")
df <- parse_all_posts(response_html)
dim(df)
```

```{r}
df %>% str
```

What does the above mean?

**🎯 ACTION POINT**: Would you organise this data frame differently?

> your text here

# Pre-processing the data

## Convert column `time` to a proper datetime type

-   Using the [lubridate package](https://lubridate.tidyverse.org/)

```{r}
library(lubridate)
```

```{r}
# Luckily the column `post_time` is already well formatted:
df %>% mutate(post_time_formatted=ymd_hms(post_time))
```

If we want to store this new column, we need to re-assign the dataframe `df`

```{r}
df <- df %>% mutate(post_time_formatted=ymd_hms(post_time))
```

```{r}
df %>% str
```

## Let's create three columns: `votes`, `answers`, `views`

-   That is a task for tidyr [pivot_wider](https://tidyr.tidyverse.org/reference/pivot_wider.html) (part of tidyverse)

```{r}
df %>% pivot_wider(id_cols=c(post_title, post_time_formatted),
                   names_from=unit,
                   values_from=number)
```

What can we improve about the DF above?

**🤝 WORKING TOGETHER**

-   Modify the functions we created in the `stackparser_questions.R` file so that:
    -   `answer` is renamed to `answers`
    -   `vote` is renamed to `votes`
-   Recreate the data frame (code below) and see the difference!
-   Your final data frame should have:
    -   No Missing Data
    -   The following columns:
        -   post_title
        -   post_time_formatted
        -   votes
        -   answers
        -   views

```{r}
response_html <- read_html("https://stackoverflow.com/questions/tagged/R")
df <- parse_all_posts(response_html)

df_formatted <- 
  df %>% 
  mutate(post_time_formatted=ymd_hms(post_time)) %>% 
  pivot_wider(id_cols=c(post_title, post_time_formatted),
              names_from=unit,
              values_from=number)

print(paste("Column names: ", paste(names(df_formatted), collapse=", ")))

df_formatted # or View(df_formatted)
```
